---
title: "Assignment 3 - CRD"
author: "COMFORT ADEGBENRO"  
format:
  html:
    code-fold: false
    embed-resources: true
    toc: true
    number-sections: false
    theme: cerulean
---

# Introduction  
This assignment will ask you to conduct a full CRD analysis, from data import through publication ready plot.  

For that, you will first need to watch (and code along with) the following CRD videos:

- CRD lecture: https://youtu.be/r5vsakgqMXc?si=8udshzlpZ-8td1f4&t=4941 (starts at 1:22:21, watch till end)

- CRD lecture pt 2, lab pt 1: https://youtu.be/GpfHxh1eK5U?si=nTle5YcB8JQklja2 
(watch whole video)  

- CRD lab pt 2: https://youtu.be/YQ1bK6uV194?si=LnDzRT69OsiMLc-r
(wathc the whole video)  

**AFTER watching the videos and developing the code in class**, then start working on this HW assignment.  

The different sections representing what we did in class were left in the script to provide you a scaffold to start from.  

# Instructions  
  - Move this script into the `code` subfolder of your `03_crd` project folder  
  - Move the file `wheat_nk_bamyan.csv` into the `data` folder of your `03_crd` project folder  
  - On this script, fill your name under the `title` section of the YAML  

# a) Setup  
Here is where we load the packages we will use.  
Make sure to load each package in the chunk below.  
```{r setup, message=FALSE}
# Loading packages
library(tidyverse)
library(car)
library(broom)
library(emmeans)
library(multcomp)

```

Now, import the data set `wheat_nk_bamyan.csv` in the chunk below.  
```{r data import}
crd_ass = read_csv("../data/wheat_nk_bamyan.csv")
head(crd_ass)

```

The **objective of this study** was to assess the effect of different N and K rates on wheat yield.  

For that, a study was conducted using a completely randomized design as the experimental design with four replicates. The treatment design was a 3 N x 3 K rate factorial, for a total of 9 treatments.  


# b) EDA tables  
Use this space to conduct exploratory data analysis of the imported data, in the form of tables.  
```{r summary}
summary(crd_ass)

```
```{r glimpse}
glimpse(crd_ass)

```


# c) Wrangling  
Here, make sure to wrangle your data set with all needed steps. Refer back to class code if you need a refresher on what was needed.  
```{r crd_asw}
crd_asw = crd_ass %>%
  mutate(rep = factor(rep),
         nrate_kgha = factor(nrate_kgha),
         krate_kgha = factor(krate_kgha)
         ) %>%
  mutate(trtname = paste0(nrate_kgha,"+",krate_kgha))

head(crd_asw)

```
```{r summary crd_asw}
summary(crd_asw)
```
## Question 1) Below each plot of this section, write a sentence explaining your visual interpretation of the plot.  

# d) EDA plots  
Now, explore the wrangled data using plots.  
Make sure to explore plots showing the main effects and the interaction.  
```{r nitrogen boxplot}
ggplot(crd_asw, aes(x= nrate_kgha,
                    y= yield_kgha,
                    color= nrate_kgha
                    )) +
  geom_boxplot() +
  geom_jitter() +
  theme(legend.position = "none")

```
The boxplot showing the effect of nitrogen rates on the yield of wheat shows that there could be nitrogen effect on yield. Specifically, the 200 kg/ha rate of nitrogen gives higher wheat yield, with the 0 kg/ha nitrogen rate having the lowest wheat yield.

```{r potassium boxplot}
ggplot(crd_asw, aes(x= krate_kgha,
                    y= yield_kgha,
                    color= krate_kgha
                    )) +
  geom_boxplot() +
  geom_jitter() +
  theme(legend.position = "none")

```
Similar to the effect of nitrogen, the highest potassium rate (60 kg/ha) seems to support more yield compared to the other rates. The 0 kg/ha of potassium resulted in the lowest yield.

```{r n by k boxplot}
ggplot(crd_asw, aes(x= nrate_kgha,
                    y= yield_kgha,
                    color= nrate_kgha
                    )) +
  geom_boxplot() +
  geom_jitter() +
  facet_grid(.~krate_kgha)
  theme(legend.position = "none")

```
The boxplot for the nitrogen by potassium shows possible interactions. For instance, as potassium rates increased from 0 to 30 to 60 kg/ha, the yield increased irrespective of the nitrogen rate. Additionally, when potassium increased from 0 to 30 kg/ha, the yield responses for the nitrogen treatments was clearly different, with the 200 kg/ha of nitrogen producing more wheat yield compared to the 100 kg/ha of nitrogen. However, at the highest level of potassium, little to no increase was observed for 200 kg/ha of nitrogen, while the yield of 100 kg/ha nitrogen increased to match up with the 200 kg/ha treatment.

# e) Statistical model  
In this section, run the appropriate model.  
Remember to set the contrast type to sum-to-zero (the default in R is set-to-zero, which is not what we want).  

Also, remember to specify all model terms needed for the treatment design.  

```{r model}
# changing to sum-to-zero contrast
options(contrasts = c("contr.sum", "contr.poly"))

# fitting the model
crd_mod = lm(yield_kgha ~ nrate_kgha * krate_kgha,
             data = crd_asw)

# summary of model
summary(crd_mod)
```

# f) ANOVA table  
Use this section to check the ANOVA table, ensuring you use type 3 sum-of-squares.
```{r ANOVA}
Anova(crd_mod, type = 3)

```
## Question 2) Based on the ANOVA table above, for which effect(s) in the ANOVA should you extract means and perform pair-wise comparisons? Explain why.  

Results from the ANOVA showed that the interaction between the two treatments (nitrogen and potassium rates) was not significant, however, the main effects of each treatment were significant. therefore, we should focus on extracting means and perform pair-wise comparisons for nitrogen, then potassium, in order to know which rates for each treatment performed better or less.

# g) Linear model assumptions  
Use this section to first compute your residuals, including studentdized residuals.  

```{r computing residuals}
crd_resid = augment(crd_mod) %>%
  mutate(.studresid = rstudent(crd_mod)) # to create the studentdized residuals

head(crd_resid)

```

## Question 3) What are the linear model assumptions that we need to check for? Create subsections (using ##) below for each of these assumptions, and write code to produce plots to check for these assumptions. For each assumption, write down your interpretation of the produced plots, and your thoughts if the assumptions hold or not.  

## Residual independence

```{r}
ggplot(crd_resid, aes(x= .fitted,
                      y= .studresid)) +
  geom_point(shape = 21,
             fill= "purple",
             size= 3,
             alpha= 0.7) +
  geom_hline(yintercept = c(-3,0,3), color = "red") +
  geom_smooth() +
  theme_bw()

```
The assumption of residual independence is satisfies as there are no patterns observed. Additionally, the smooth line sits on the th zero of the y axis and the shaded areas are around the zero mark too.

## Residual variance homogeneity

```{r}
ggplot(crd_resid, aes(x= .fitted,
                      y= .studresid)) +
  geom_point(shape = 21,
             fill= "purple",
             size= 3,
             alpha= 0.7) +
  geom_hline(yintercept = c(-3,0,3), color = "red") +
  geom_smooth() +
  theme_bw()

```
The residuals variance are homogeneous as the spread are uniform across the plot. There is no distinct pattern different from other parts of the plot. There is no heteroscedasticity

## Residual normality

```{r qq plot}
ggplot(crd_resid, aes(sample = .studresid)) +
         stat_qq() +
         stat_qq_line() +
  theme_bw()
```
The qq-plot shows that some points are slighly deviated from the line, especially in the middle. It is not on the extreme end, so can be accepted.

```{r density plot}
ggplot(crd_resid, aes(x= .studresid)) +
  geom_density() +
  scale_x_continuous(breaks = c(-2,0,2), limits = c(-3,3)) +
  theme_bw()
```
The density plots looks like it has two peaks with the main peak close to the mean. To confirm if the rule of normal distribution is met, more tests on the normal distribution and spread of the residuals are carried out below.

```{r Extra tests for heteroscedasticity}

# Shapiro-Wilk Test
shapiro.test(residuals(crd_mod)) # If p < 0.05, residuals significantly deviate from normality

# Breusch-Pagan Test
library(lmtest)
bptest(crd_mod)  # If p < 0.05, heteroscedasticity is present

# Non-Constant Variance Test
ncvTest(crd_mod)  # If p > 0.05, no significant heteroscedasticity detected

# White’s Test for Non-Linear Heteroscedasticity
bptest(crd_mod, ~ fitted(crd_mod) + 
         I(fitted(crd_mod)^2))  # p > 0.05 means no evidence of heteroscedasticity.

```
All p values were above 0.05, hence the assumption is met for normal distribution and doublecheck for homoscedasticity

## Outlier detection

```{r}
ggplot(crd_resid, aes(x= .fitted,
                      y= .studresid)) +
  geom_point(shape = 21,
             fill= "purple",
             size= 3,
             alpha= 0.7) +
  geom_hline(yintercept = c(-3,0,3), color = "red") +
  geom_smooth()
```
All residuals are within the -3 and 3 range, so there are no outliers. The assumption for outliers is met.

# h) Model means  
Use this section to extract model means according to your decision of the ANOVA table effects. 
```{r emmeans for nrate only}
crd_means_n = emmeans(crd_mod, ~nrate_kgha)
crd_means_n

```
```{r emmeans for krate only}
crd_means_k = emmeans(crd_mod, ~krate_kgha)
crd_means_k

```

# i) Pairwise comparisons  
Use this section to perform pair-wise comparison of the means extracted above.  

Make sure to reverse the order from high to low, use letters instead of numbers, and no adjustment. 


```{r pairwise comparison n}
crd_cld_n = cld(crd_means_n,
                reversed = T,
                Letters = letters,
                adjust = "none") %>%
  as_data_frame() %>% # to convert pwc output to dataframe
  mutate(letter = trimws(.group)) # to add an extra column removing space in the letters

crd_cld_n
```


```{r pairwise comparison k}
crd_cld_k = cld(crd_means_k,
                reversed = T,
                Letters = letters,
                adjust = "none") %>%
  as_data_frame() %>% # to convert pwc output to dataframe
  mutate(letter = trimws(.group)) # to add an extra column removing space in the letters

crd_cld_k
```

# g) Final plot(s)  
Produce final plot(s) to show your results.  
Make sure to include both **raw data** and **statistical model summary (i.e., letter separation)**.  

Only plots for main effects for nitrogen and potassium are shown, since there was no significant interactive effect of both treatment factors on wheat yield. 

```{r boxplot for nitrogen}
ggplot(data= crd_asw,
               aes(x= nrate_kgha, 
                   y= yield_kgha,
                   color= nrate_kgha)) +
  geom_boxplot(width = 0.4) +
  geom_jitter() +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  geom_label(data = crd_cld_n,
             aes(x= nrate_kgha, 
                 y= emmean,
                 label = letter)) +
  theme(legend.position = "none",
        axis.line = element_line(linewidth = 0.6),
        axis.text = element_text(face = "bold", size= 9, 
                                 color = "black"),
        axis.title = element_text(face = "bold"),
        axis.ticks = element_line(linewidth = 0.8)) +
  labs(x= "Nitrogen Rate (kg/ha)", 
       y= "Wheat Yield (kg/ha)")
ggsave("nplot.png", path = "../output/")

```
As seen in the figure above, increasing nitrogen rate from 0 kg/ha to 100 kg/ha significantly increased wheat yield. However, increasing nitrogen rate up to 200 kg/ha did not result in additional increase in yield. Hence, 100 kg/ha of nitrogen is recommended to maximize wheat yield.

```{r boxplot for potassium}
ggplot(data= crd_asw,
               aes(x= krate_kgha, 
                   y= yield_kgha,
                   color= krate_kgha)) +
  geom_boxplot(width = 0.4) +
  geom_jitter() +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  geom_label(data = crd_cld_k,
             aes(x= krate_kgha, 
                 y= emmean,
                 label = letter)) +
  theme(legend.position = "none",
        axis.line = element_line(linewidth = 0.6),
        axis.text = element_text(face = "bold", size= 9, 
                                 color = "black"),
        axis.title = element_text(face = "bold"),
        axis.ticks = element_line(linewidth = 0.8)) +
  labs(x= "Potassium Rate (kg/ha)", 
       y= "Wheat Yield (kg/ha)")
ggsave("kplot.png", path = "../output/")
```
Potassium rates of 30 and 60 kg/ha resulted in the highest values of wheat yield, and there was no statistical difference between these two potassium rates. To maximize yield, a 30 kg/ha potassium rate is recommended.

Be mindful of your plot to ensure you are creating it in an intuitive way for your audience in relation to interpreting the letter separation structure you used.  

In class, we did not have time to make it publication-ready (e.g., a plot looking professional and ready to be included in a manuscript).

Here, make sure you deliver final plot(s) that is/are publication ready.  

Make sure to provide an interpretation for the plot(s) created, including which treatment levels were significantly higher/lower.


# Submitting your work  
Once you have developed all the code and answers, make sure to Render this quarto file.  

**Notes on rendering**:  

- Make sure to render your work and inspect how the final html look like.  
- If it does not look professional for whatever reason, then fix the issue, re-render it, recheck.  
- Only send me your work once your html file looks professional.  
  - **DO NOT** delete the file's heading levels (# and ##). They set up the proper heading 1 and 2 levels, and I use them to guide my grading.  
  - If a given chunk is also outputting warnings or messages, inhibit this behavior by changing the chunk options `message` and `warning` to `FALSE` (I'll demo this in a moment).  
  
  - If, after rendered, 2 lines of text are connected and you wish to "break line" between them, add 2 extra spaces after the first one (I'lld demo this in a moment).  

After rendering, an .html file will be created on your `code` folder.  

Rename this file to `Assignment-03-crd-LASTNAME.html`.    

For ex., mine would be `Assignment-03-crd-Bastos.html`.

Submit your work by uploading the **html** file to **eLC Assigment #3 - CRD** by Feb 3rd 11:59 pm.  
  




